Imperative Programming with Variational Effects

### 2018-08-08 paper/paper.tex
work on related work section

>>>>

\chapter{Related Work}
\label{ch:rw}

As discussed at various points in this thesis, the formal semantics we have developed for variational imperative languages
is in part a synthesis of existing techniques developed for existing variational interpreters. The present work builds upon this
by bringing all of these techniques together and providing a formal semantics for reasoning about them. It also demonstrates
how to support catchable exceptions in a variational environment, which to our knowledge is absent from previous work.
In this chapter we compare our work to preceding work on variational execution, highlighting the development and implementation
of several techniques utilized here.

 K\"astner et al. \cite{varwhile} developed a proof-of-concept variational interpreter for the \textsc{While} language. This works shares a
 significant amount in common with our work on IMP, as both utilize toy imperative languages to develop a theory of variational execution
 with side effects. This work originated the concepts of maintaining a variational context during execution, defining a variational store as
 a variational map with lookup and update operations that take the variational context as an argument, and the use of the $\mathit{whenTrue}()$
 helper function in conjunction with control structures. However, the variational \textsc{While} language only supports a single type of side
 effect---mutable state---and therefore is incapable of analyzing the interactions between multiple different types of effects and variation.
 Additionally, the evaluation of the language is only presented as Scala functions rather than a formal big-step semantics.

 Around the same time, Kim, Khurshid, and Batory developed \emph{shared execution} for testing software product lines \cite{sharedexec}.
 Shared execution primarily focuses on low-level support for variation at the level of a language VM.
 Unlike the variational \textsc{While} language, shared execution is not built on the formal foundation of the choice calculus, and thus its
 support for variation is handled on a more ad-hoc case-by-case basis, in particular in its merging of shared contexts. Nevertheless,
 shared execution does contain an analogue of the variational context with variational store pattern for dealing with variational mutable state.

 Another seemingly independent development of variational execution was Austin and Flanagan's \emph{faceted evaluation} for Javascript \cite{faceted}.
 In this case, choices, or \emph{facets} as they are termed by Austin and Flanagan, represent desirable security properties of Javascript, where
 private data lows correspond to the right alternative and public dataflows correspond to the left alternative.
 Again, this work contains an analogue of the variational context with variational store pattern, in this case referring to the variational context as
 the \emph{program counter}. Faceted evaluation only allows dimensions as tags for its choices, rather than arbitrary boolean expressions, which
 decreases the expressiveness of the language but obviates the need for satisfiability checking. Faceted evaluation additionally provides a formal
 semantics for a language they call $\lambda^{\mathit{facet}}$. It is also the only approach that attempts to provide a mechanism for variational IO
 side effects such as interacting with the file system. However, in comparison with the other approaches to variational execution presented here,
 faceted evaluation is primarily concerned with dataflow security and not with developing general models of variational programming.

 Building on the above work with the \textsc{While} language, Nguyen, K\"astner, and Nguyen developed the Varex interpreter for a subset of PHP \cite{varex}.
 Varex successfully demonstrated the ability to test plugin interactions in the highly configurable WordPress application. Varex adds support for throwing
 exceptions, but they are not catchable. Varex's exceptions are also reported directly when encountered as a side effect of execution, rather than producing
 a variational error value as the result of execution. Nevertheless, Varex demonstrates the utility of developing a variational interpreter for a subset of an
 existing language.

\chapter{Conclusion and Future Work}
\label{ch:conc}
